---
title: "Subgroup vs Unaffected Differential Gene Expression (DGE) Analysis"
author: "Shaurita D. Hutchins"
date: "`r format(Sys.Date(), 'Last updated on %Y-%m-%d')`"
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
code_download: true
editor_options:
  chunk_output_type: console
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

The [subsetted main analysis](subsetted-condition-analysis.html) tests Affected vs Unaffected in one model. This page repeats that workflow for each genetic subgroup vs Unaffected so subgroup means “samples carrying that label in `Subgroup`” compared to all Unaffected samples, not subgroup vs subgroup.

We analyze Potassium, HPP, Ion_Channel, CHRN, Mito, and RBC. The statistical model is `~ Batch + group` where `group` is subgroup vs Unaffected. VST stabilizes variance for visualization; `removeBatchEffect()` removes batch from the VST matrix for heatmaps only; Wald tests and p-values still come from `DESeq()` on raw counts, same as the main analysis.

We save tables and figures for every subgroup under `output/subgroup-{name}-analysis/`. This HTML shows only Mito and RBC padj &lt; 0.05 heatmaps and the RBC volcano (supplemental figures in the paper). Objects are redrawn in the document so the site does not depend on fragile image paths.

# DGE Analysis Setup

Activate `renv` and `source()` the same `code/helpers.R` as the subsetted analysis so helper names match.


```{r, load libraries, warning = FALSE, message = FALSE}
library(tidyverse) # Available via CRAN
library(DESeq2) # Available via Bioconductor
library(RColorBrewer) # Available via CRAN
library(pheatmap) # Available via CRAN
library(genefilter) # Available via Bioconductor
library(limma) # Available via Bioconductor
library(gprofiler2) # Available via CRAN
library(biomaRt) # Available via Bioconductor
library(plotly) # Available via CRAN
library(ggpubr) # Available via CRAN
library(rmarkdown) # Available via CRAN
library(clusterProfiler) # Available via Bioconductor
library(org.Hs.eg.db) # Available via Bioconductor
library(ggrepel) # Available via CRAN
library(ReactomePA) # Available via Bioconductor
library(mygene) # Available via Bioconductor
library(DOSE) # Available via Bioconductor
library(enrichR) # Available via Bioconductor
library(STRINGdb) # Available via Bioconductor
library(EnhancedVolcano)
library(ComplexHeatmap)
```

```{r, include=FALSE}
source("code/helpers.R", local = knitr::knit_global())
```

## Data Import

Counts and metadata match the main analysis. `Subgroup` can list multiple labels per sample (comma-separated); we split and treat each label as its own subgroup list. The chunk loads everything once; `genes_biomart` avoids repeated Ensembl lookups inside the loop.

```{r, import data}
# Load full metadata and counts (no subgroup filter yet)
sample_metadata_full <- read_csv("data/Metadata_2024_11_20.csv")
row.names(sample_metadata_full) <- sample_metadata_full$ID
counts_full <- read_tsv("data/star-salmon/salmon_merged_gene_counts_length_scaled.tsv")

# Load genes of interest
genes_of_interest <- read_csv("data/Prioritized_Genes_From_WGS_2024_11_19.csv") %>%
  pull(Genes) %>%
  unique() %>%
  na.omit()
genes_of_interest <- unique(c(genes_of_interest, "CHI3L1"))

# Process counts: use first column for row names, keep gene info
counts_full <- data.frame(counts_full, row.names = 1, check.names = FALSE)
counts_full$Ensembl_ID <- row.names(counts_full)
drop <- c("Ensembl_ID", "gene_name")
gene_info <- counts_full[, drop]
counts_full <- counts_full[, !(names(counts_full) %in% drop)]

# Derive list of subgroups from metadata (exclude Unaffected)
subgroup_list_raw <- unique(unlist(strsplit(trimws(sample_metadata_full$Subgroup), ",\\s*")))
subgroups_to_analyze <- setdiff(subgroup_list_raw, c("Unaffected", ""))
subgroups_to_analyze <- subgroups_to_analyze[!is.na(subgroups_to_analyze)]

# Subgroups for which we show heatmaps and volcano in the report
heatmap_display_subgroups <- c("Mito", "RBC")
volcano_display_subgroup <- "RBC"

# Retrieve gene annotations once (shared across subgroups)
genes_biomart <- retrieve_gene_info(
  values = gene_info$Ensembl_ID,
  filters = "ensembl_gene_id_version"
)
```

## Annotation colors

`ann_colors` maps metadata columns to colors so heatmap legends match the publication styling.

```{r, heatmap-colors, out.height="570px", out.width ="800px"}
study_group_colors <- c(
  "ENE" = "#BFDFBF",
  "IMM" = "#EEBFEE",
  "SOL" = "#FFFF33",
  "STR" = "#BFBFFF",
  "RBC" = "#FFD700",
  "N/A" = "darkgray"
)

group_colors <- c(
  "Potassium" = "#80B1D3",
  "HPP" = "#FB8072",
  "Ion_Channel" = "#FDB462",
  "CHRN" = "#B3DE69",
  "RBC" = "#BC80BD",
  "Mito" = "#CCEBC5",
  "Unaffected" = "gray40"
)

ann_colors <- list(
  Batch = c(B1 = "purple", B2 = "firebrick", B3 = "goldenrod"),
  Affected = c(Affected = "green4", Unaffected = "navy"),
  Study_group = study_group_colors,
  group = group_colors
)
```

`draw_subgroup_heatmap()` wraps `ComplexHeatmap::pheatmap()` so the analysis loop and the report use identical colors and layout.

```{r, heatmap-helper, include=FALSE}
# Keep the heatmap drawing code in one place so the loop and report sections
# render the same plot settings without relying on saved PNG paths.
draw_subgroup_heatmap <- function(heatmap_matrix, annotation_col) {
  ComplexHeatmap::draw(
    ComplexHeatmap::pheatmap(
      heatmap_matrix,
      color = colorRampPalette(rev(brewer.pal(n = 9, name = "RdYlBu")))(50),
      annotation_col = annotation_col,
      annotation_colors = ann_colors,
      fontsize = 7,
      angle_col = "0",
      cellheight = 10,
      cellwidth = 24,
      border_color = "black",
      heatmap_legend_param = list(title = "")
    )
  )
}
```

## Run DGE for all subgroups

The loop runs one complete DGE run per subgroup. Full rank means every `Batch` × `group` combination needed for `~ Batch + group` is estimable; if not (often too few samples in a batch), we skip with a message. Outputs go to `output/subgroup-{name}-analysis/`.

Only RBC gets a saved volcano PNG. Mito, RBC padj &lt; 0.05 heatmaps, and the RBC volcano ggplot are also held in memory for the sections below.

```{r, run-all-subgroups, results="hold", fig.width=14, fig.height=10, out.width="100%"}
subgroup_report_heatmaps <- list()
subgroup_report_volcano <- NULL

for (subgroup_name in subgroups_to_analyze) {
  subgroup_lower <- tolower(subgroup_name)
  outpath_var <- paste0("subgroup-", subgroup_lower, "-analysis")
  out_dir <- file.path("output", outpath_var)
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

  # Filter metadata to this subgroup + Unaffected
  sample_metadata <- sample_metadata_full %>%
    mutate(Subgroup_list = strsplit(Subgroup, ",")) %>%
    mutate(in_subgroup = purrr::map_lgl(Subgroup_list, ~ subgroup_name %in% trimws(.))) %>%
    mutate(group = dplyr::case_when(
      in_subgroup ~ subgroup_name,
      Subgroup == "Unaffected" ~ "Unaffected",
      TRUE ~ NA_character_
    )) %>%
    filter(!is.na(group))

  n_sub <- sum(sample_metadata$group == subgroup_name)
  n_unaff <- sum(sample_metadata$group == "Unaffected")
  if (n_sub < 1) {
    message("Skipping ", subgroup_name, ": no samples in subgroup.")
    next
  }

  # Subset counts to these samples and match order
  sample_metadata$ID <- as.character(sample_metadata$ID)
  counts <- counts_full[, colnames(counts_full) %in% sample_metadata$ID, drop = FALSE]
  counts <- counts[, match(sample_metadata$ID, colnames(counts))]
  stopifnot(all(colnames(counts) == sample_metadata$ID))

  # DESeq2
  sample_metadata <- sample_metadata %>%
    mutate(
      Family = factor(Family),
      Affected = factor(Affected),
      Batch = factor(Batch),
      Sex = factor(Sex),
      Ancestry = factor(Ancestry),
      Category = factor(Category),
      group = factor(group)
    )

  # Skip if design matrix is not full rank (e.g. only one batch in this subgroup)
  mm <- model.matrix(~ Batch + group, sample_metadata)
  if (qr(mm)$rank < ncol(mm)) {
    message("Skipping ", subgroup_name, ": model matrix not full rank (e.g. batch and group confounded).")
    next
  }

  dds <- DESeqDataSetFromMatrix(
    countData = round(counts),
    colData = sample_metadata,
    design = ~ Batch + group
  )
  keep <- rowSums(counts(dds)) >= 50
  dds <- dds[keep, ]
  dds <- DESeq(dds)

  # VST and limma batch correction
  vsd <- vst(dds, blind = FALSE)
  counts_vst <- assay(vsd)
  write.csv(counts_vst, file.path(out_dir, "counts_vst.csv"))
  mm <- model.matrix(~ Batch + group, colData(vsd))
  counts_vst_limma <- limma::removeBatchEffect(counts_vst, batch = vsd$Batch, design = mm)
  write.csv(counts_vst_limma, file.path(out_dir, "counts_vst_limma.csv"))
  vsd_limma <- vsd
  assay(vsd_limma) <- counts_vst_limma

  # Results
  res_subgroup_vs_unaff <- results(dds, contrast = c("group", subgroup_name, "Unaffected"))
  res_file_name <- file.path(out_dir, paste0("res_", subgroup_lower, "_vs_unaff.csv"))
  res_subgroup_vs_unaff_df <- process_and_save_results(res_subgroup_vs_unaff, res_file_name)
  res_subgroup_vs_unaff_df <- arrange(res_subgroup_vs_unaff_df, padj)

  res_subgroup_vs_unaff_df_05 <- subset(res_subgroup_vs_unaff_df, padj < 0.05)
  res_subgroup_vs_unaff_df_1 <- subset(res_subgroup_vs_unaff_df, padj < 0.1)
  res_subgroup_vs_unaff_df_padj1_lfc1 <- subset(res_subgroup_vs_unaff_df_1, abs(log2FoldChange) >= 1)
  res_subgroup_vs_unaff_df_padj05_lfc1 <- subset(res_subgroup_vs_unaff_df_05, abs(log2FoldChange) >= 1)

  # Annotated results with gene names
  res_subgroup_vs_unaff_df_genename <- res_subgroup_vs_unaff_df
  res_subgroup_vs_unaff_df_genename$Ensembl_ID <- row.names(res_subgroup_vs_unaff_df)
  res_subgroup_vs_unaff_df_genename <- merge(
    res_subgroup_vs_unaff_df_genename,
    gene_info,
    by = "Ensembl_ID",
    all.x = TRUE
  )
  res_subgroup_vs_unaff_df_genename <- res_subgroup_vs_unaff_df_genename[
    , c("gene_name", "Ensembl_ID", setdiff(names(res_subgroup_vs_unaff_df_genename), c("gene_name", "Ensembl_ID")))
  ]
  res_subgroup_vs_unaff_df_genename <- res_subgroup_vs_unaff_df_genename[
    order(res_subgroup_vs_unaff_df_genename$padj),
  ]
  write.csv(
    res_subgroup_vs_unaff_df_genename,
    file.path(out_dir, "res_subgroup_vs_unaff_df_genename.csv"),
    row.names = FALSE
  )

  # Filtered result CSVs
  process_and_save_filtered_results(
    res_subgroup_vs_unaff_df_genename,
    padj_threshold = 0.05, lfc_threshold = NULL,
    outpath = outpath_var,
    filename = paste0("res_", subgroup_lower, "_vs_unaff_df_genename_05.csv")
  )
  process_and_save_filtered_results(
    res_subgroup_vs_unaff_df_genename,
    padj_threshold = 0.05, lfc_threshold = 1,
    outpath = outpath_var,
    filename = paste0("res_", subgroup_lower, "_vs_unaff_df_genename_padj05_lfc1.csv")
  )
  process_and_save_filtered_results(
    res_subgroup_vs_unaff_df_genename,
    padj_threshold = 0.1, lfc_threshold = NULL,
    outpath = outpath_var,
    filename = paste0("res_", subgroup_lower, "_vs_unaff_df_genename_padj1.csv")
  )
  process_and_save_filtered_results(
    res_subgroup_vs_unaff_df_genename,
    padj_threshold = 0.1, lfc_threshold = 1,
    outpath = outpath_var,
    filename = paste0("res_", subgroup_lower, "_vs_unaff_df_genename_padj1_lfc1.csv")
  )

  # Heatmaps: build matrix and save PNG for every subgroup
  res_subgroup_vs_unaff_df <- res_subgroup_vs_unaff_df[order(res_subgroup_vs_unaff_df$padj), ]
  ensembl_to_gene <- setNames(gene_info$gene_name, gene_info$Ensembl_ID)

  # Heatmap padj < 0.1, |log2FC| >= 1
  topgenes_byensemblid <- rownames(res_subgroup_vs_unaff_df_padj1_lfc1)
  if (length(topgenes_byensemblid) > 0) {
    topgenes_mat <- assay(vsd_limma)[topgenes_byensemblid, , drop = FALSE]
    if (nrow(topgenes_mat) == 1) {
      topgenes_mat <- topgenes_mat - mean(topgenes_mat)
    } else {
      topgenes_mat <- topgenes_mat - rowMeans(topgenes_mat)
    }
    rownames(topgenes_mat) <- ensembl_to_gene[rownames(topgenes_mat)]
    topgenes_mat <- topgenes_mat[order(res_subgroup_vs_unaff_df$padj[topgenes_byensemblid]), , drop = FALSE]
    annotation_df <- as.data.frame(colData(vsd_limma)[, c("group", "Affected"), drop = FALSE])
    png(file.path(out_dir, paste0(subgroup_name, "_padj1_heatmap.png")), width = 14, height = 10, units = "in", res = 1200)
    draw_subgroup_heatmap(topgenes_mat, annotation_df)
    dev.off()
  }

  # Heatmap padj < 0.05, |log2FC| >= 1
  topgenes_byensemblid_05 <- rownames(res_subgroup_vs_unaff_df_padj05_lfc1)
  if (length(topgenes_byensemblid_05) > 0) {
    topgenes_mat_05 <- assay(vsd_limma)[topgenes_byensemblid_05, , drop = FALSE]
    if (nrow(topgenes_mat_05) == 1) {
      topgenes_mat_05 <- topgenes_mat_05 - mean(topgenes_mat_05)
    } else {
      topgenes_mat_05 <- topgenes_mat_05 - rowMeans(topgenes_mat_05)
    }
    rownames(topgenes_mat_05) <- ensembl_to_gene[rownames(topgenes_mat_05)]
    topgenes_mat_05 <- topgenes_mat_05[order(res_subgroup_vs_unaff_df$padj[topgenes_byensemblid_05]), , drop = FALSE]
    df_ann <- as.data.frame(colData(vsd_limma)[, "group", drop = FALSE])
    if (subgroup_name %in% heatmap_display_subgroups) {
      # Store the heatmap inputs so the report can redraw the same figure inline.
      subgroup_report_heatmaps[[subgroup_name]] <- list(
        mat = topgenes_mat_05,
        annotation_col = df_ann
      )
    }
    png(file.path(out_dir, paste0(subgroup_name, "_significant_padj05_heatmap.png")), width = 10, height = 8, units = "in", res = 300)
    draw_subgroup_heatmap(topgenes_mat_05, df_ann)
    dev.off()
  }

  # Volcano: save only for RBC so it is not overwritten
  if (subgroup_name == volcano_display_subgroup) {
    output_volcano_file <- file.path(out_dir, paste0("volcano-plot-", subgroup_lower, ".png"))
    subgroup_report_volcano <- generate_volcano_plot(
      res_data = res_subgroup_vs_unaff_df_genename,
      gene_labels = "gene_name",
      x_col = "log2FoldChange",
      y_col = "padj",
      select_genes = "",
      xlab_text = bquote(~ Log[2] ~ "fold change"),
      ylab_text = bquote(~ -Log[10] ~ "padj"),
      p_cutoff = 0.05,
      fc_cutoff = 1.0,
      xlim_range = c(-25, 25),
      ylim_range = c(0, 7),
      output_file = output_volcano_file,
      print_plot = FALSE
    )
  }

  # MA-plot significant genes CSV
  maplot <- generate_ma_plot(
    data = res_subgroup_vs_unaff_df,
    genenames = as.vector(res_subgroup_vs_unaff_df_genename$gene_name),
    main_title = paste0(subgroup_name, " vs Unaffected MA Plot"),
    top_genes = 30
  )
  significant_data <- maplot$data %>%
    filter(grepl("Up|Down", sig)) %>%
    mutate(direction = ifelse(grepl("Up", sig), "Up", "Down")) %>%
    dplyr::select(-sig)
  write.csv(
    significant_data,
    file.path(out_dir, paste0("res_", subgroup_lower, "_vs_unaff_siggenes.csv")),
    row.names = FALSE
  )

  # Save RData for this subgroup (for optional reload)
  save(
    dds, vsd, vsd_limma,
    res_subgroup_vs_unaff_df, res_subgroup_vs_unaff_df_genename,
    res_subgroup_vs_unaff_df_padj05_lfc1, res_subgroup_vs_unaff_df_padj1_lfc1,
    file = file.path(out_dir, paste0(subgroup_name, "-subgroup-analysis.RData"))
  )

  message("Processed: ", subgroup_name, " (n = ", n_sub, " vs Unaffected n = ", n_unaff, ")")
}
```

## Heatmaps (Mito and RBC only)

Every subgroup gets padj &lt; 0.1 and padj &lt; 0.05 heatmaps on disk. Here, Mito and RBC padj &lt; 0.05 panels reuse the batch-corrected VST matrix from the loop, row-mean centered so color shows deviation from each gene’s average across displayed samples. Columns are annotated with `group` (subgroup vs Unaffected).

### Mito vs Unaffected

Genes pass padj &lt; 0.05 and |log2FC| ≥ 1 for Mito vs Unaffected in this run.

```{r, heatmap-mito-padj05, fig.width=10, fig.height=8, echo=FALSE}
hm_mito <- subgroup_report_heatmaps[["Mito"]]
if (!is.null(hm_mito)) {
  draw_subgroup_heatmap(hm_mito$mat, hm_mito$annotation_col)
} else {
  message("No Mito padj<0.05 heatmap in this run (no significant genes or subgroup skipped).")
}
```

### RBC vs Unaffected

Same cutoffs for RBC vs Unaffected.

```{r, heatmap-rbc-padj05, fig.width=10, fig.height=8, echo=FALSE}
hm_rbc <- subgroup_report_heatmaps[["RBC"]]
if (!is.null(hm_rbc)) {
  draw_subgroup_heatmap(hm_rbc$mat, hm_rbc$annotation_col)
} else {
  message("No RBC padj<0.05 heatmap in this run (no significant genes or subgroup skipped).")
}
```

## Volcano plot (RBC only)

The volcano plot shows log2 fold change (RBC vs Unaffected) on the x-axis and −log10 FDR (`padj`) on the y-axis. Points above the horizontal line pass padj &lt; 0.05, and vertical lines mark |log2FC| = 1. The figure is the ggplot from the loop; a copy is still written to `output/subgroup-rbc-analysis/volcano-plot-rbc.png`.

```{r, volcano-rbc, fig.width=14, fig.height=8, echo=FALSE}
if (!is.null(subgroup_report_volcano)) {
  print(subgroup_report_volcano)
} else {
  message("RBC volcano plot not available (RBC subgroup skipped or not in subgroups_to_analyze).")
}
```

## Output summary

For each subgroup, the following were written under `output/subgroup-{name}-analysis/` for reproducibility and sharing:

- `counts_vst.csv`, `counts_vst_limma.csv`
- `res_{name}_vs_unaff.csv`, `res_subgroup_vs_unaff_df_genename.csv`
- Filtered result CSVs (padj 0.05 / 0.1, with and without LFC threshold)
- `{Subgroup}_padj1_heatmap.png`, `{Subgroup}_significant_padj05_heatmap.png`
- `res_{name}_vs_unaff_siggenes.csv` (MA-plot significant genes)
- For RBC only: `volcano-plot-rbc.png`
- `{Subgroup}-subgroup-analysis.RData` (DESeq2 and result objects for reproducibility)
